Prediction of Protein-Protein Interactions on the Human and Rice Interactomes

### 2020-09-11 nicolas/PPIPL3/ReportV2.tex

\usepackage{babel}
\begin{document}
\title{Prediction of Protein-Protein Interactions on the Human and Rice Interactome }
\author{Nicolás Antonio López Rozo}
\maketitle
\begin{abstract}

Previous Network-based efforts to predict unmapped protein-protein
interactions (PPI's) suggest that proteins with multiple paths of
length 3 (L3) are more likely to be connected. This paper extends
this so-called L3 principle by taking into account feature extraction
and using \texttt{XGBoost} techniques for prediction. In particular,
we train the model using handcrafted features as well as features
learned from embeddings using \texttt{Node2Vec}. Our main result shows
that while L3 remains an important principle for predicting links,
the approach is outperform by using embedded features. The mentioned
approaches are compared using the human and the rice interactomes.
\end{abstract}
\begin{comment}
need include a feature that compares the dimension of the embedding
is useful for prediction with L3 to show that they independent measures
- we do not want to be learning L3 through node2vec
\end{comment}

@@

Proteins are the chief actors of biological functionality inside of
the cell, and as such, do not usually work as single substances but
rather as a part of dynamic networks, interacting with other substances
for a wide variety of purposes \cite{Lin2017}.

Protein-protein interactions (PPIs) have a key role in a variety of
biological processes such as signal transduction, homeostasis control,
stress responses, plant defense and organ formation. On the molecular
level, PPIs have importance in protein phosphorylation, transcriptional
co-factor recruitment, transporter activation and many others, thus
playing an essential role in many physiological and developmental
processes in virtually all organisms \cite{Zhang_2010_PPI}.

Prediction of potentially relevant, yet unexplored PPIs is a current
research topic on bioinformatics and therefore, several authors have
proposed different methods for extrapolating information from the
existing PPI networks. As presented by Kovacs et al (2019), information
related to counting paths of length 3 (L3) seems to outperform state-of-the-art
methods when it comes to predict interactions among proteins for a
variety of model organisms such as yeast (\emph{S. cerevisiae}), Arabidopsis
(\emph{A. thaliana}), worm (\emph{C. elegans}), fly (\emph{D. melanogaster}),
fission yeast (\emph{S. pombe}) and mouse (\emph{M. musculus}), as
well as on the human interactome \cite{Kovacs2019}.

The most common way to validate PPI's is the \emph{Yeast-Two-Hybrid}
technique (also known as \emph{two-hybrid screening} or \emph{Y2H}),
which is based on the expression of a specific reporter gene that
activates by the binding of a DNA-binding Domain (DB) and an Activation
Domain (AD) of a Transcription Factor that binds to an Upstream Activation
Sequence (UAS). For the Y2H technique, a protein is fused to the DB
domain (known as \emph{bait}) and another one to the AD (known as
\emph{prey}). If the proteins do not interact, then the reporter gene
is not expressed. Otherwise, the reporter gene expression is activated
by the activation domain.

\begin{comment}
NL: I will make this figure myself
\end{comment}

@@

Having several of these Y2H results allows scientists to establish
a PPI network, where all known interactions for each protein are represented.
Several algorithms are proposed over these networks in order to predict
unknown interactions. In this report, three prediction methods are
presented and their results are shown: Common Neighbors (\textbf{CN}),
which uses the length-2 path count; raw count of paths of length 3
(\textbf{A3}); and degree-normalized length-3 paths score (\textbf{L3}).

Focus of the present study is to evaluate different methods for predicting
protein-protein interactions (PPIs) using the existing knowledge of
the network, which is an undirected graph. The traditional way is
usually based on social networks analysis, more specifically on the
Triadic Closure Principle (TCP), that states that the more common
shared friends that two people have, the more likely that they know
each other. As shown by previous studies, the mentioned approach fails
because it does not consider the structural and chemical properties
of the proteins \cite{Kovacs2019}.

@@

\section{Materials and Methods}

@@

\begin{figure}[h]
\caption{Methods Comparison for \emph{HI-II-14}}

\noindent \centering{}\includegraphics[width=1\columnwidth]{\string"/home/ocin/Documents/02 Asignaturas/10 GraphAnalytics/Project1/hi-ii-14.txt\string".eps}
\end{figure}

\begin{figure}[h]
\caption{Methods Comparison for \emph{HI-TESTED}}

\noindent \centering{}\includegraphics[width=1\columnwidth]{\string"/home/ocin/Documents/02 Asignaturas/10 GraphAnalytics/Project1/hi-tested.txt\string".eps}
\end{figure}

>>>>

\usepackage{babel}
\begin{document}
\title{Prediction of Protein-Protein Interactions on the Human and Rice Interactomes}
\author{Nicolás Antonio López Rozo}
\maketitle
\begin{abstract}

Network analysis for  predict unmapped protein-protein interactions (PPI) suggest that the higher the number of paths of
length~3 (L3) between two proteins, the more likely they are to interact. This paper extends previous work based on the L3 principle by taking into account the representation learning of node features of the PPI network. In particular, we train an XGBoost model using L3 and handcrafted features, as well as embeddings from \texttt{Node2Vec}. Our main result shows that while L3 is an important feature for predicting links, best performance is achieved when combined with edge embeddings. The proposed approached is evaluated for the human and rice interactomes.
\end{abstract}
\begin{comment}
need include a feature that compares the dimension of the embedding
is useful for prediction with L3 to show that they independent measures
- we do not want to be learning L3 through node2vec
\end{comment}

@@

Proteins are the key actors of biological processes inside cells. As they carry out a variety of tasks, cells do not work as single agents but
rather as a part of dynamic networks of protein-protein interaction (PPI)~\cite{Lin2017}. PPI networks play a key role in a variety of interdependent mechanisms, including signal transduction, homeostasis control, stress responses, plant defense and organ formation. At the molecular level, PPIs play an essential role in many physiological and developmental processes, including protein phosphorylation, transcriptional co-factor recruitment, transporter activation~\cite{Zhang_2010_PPI}.

Prediction of potentially relevant yet unexplored PPIs is a current research topic on bioinformatics. Several authors have
proposed different methods for extrapolating information from the existing PPI networks. Kovacs et al. (2019) introduces an approach which relies on counting paths of length 3 (L3) to predict interactions among proteins for a variety of model organisms. The proposed approach outperform previous methods in different PPI networks of yeast (\emph{S. cerevisiae}), Arabidopsis (\emph{A. thaliana}), worm (\emph{C. elegans}), fly (\emph{D. melanogaster}), fission yeast (\emph{S. pombe}) and mouse (\emph{M. musculus}), as well as for the human interactome \cite{Kovacs2019}.

A most common way to uncover (or validate) PPIs is the \emph{Yeast-Two-Hybrid} technique (also known as \emph{two-hybrid screening} or \emph{Y2H}),
which is based on (i) the expression of a specific reporter gene that is activated by the binding of a DNA-binding Domain (DB); and (ii) an activation domain (AD) of a transcription factor that binds to an upstream activation sequence (UAS). For the Y2H technique, a protein is fused to the DB domain (known as \emph{bait}) and another one to the AD (known as \emph{prey}). If the proteins do not interact, then the reporter gene
is not expressed. Otherwise, the reporter gene expression is activated by the activation domain.

\begin{comment}
NL: I will make this figure myself
JF: Good idea
\end{comment}

@@

Based on the outcome of numerous Y2H experiments PPI networks can be constructed, in which known interactions for each protein are represented. Several algorithms are proposed over these networks in order to predict yet unknown interactions. This paper evaluates predictions based on three measures: Common Neighbors (\textbf{CN}), based on the count of paths of length 2; the raw count of paths of length 3
(\textbf{A3}); and the degree-normalized count of paths of length 3(\textbf{L3}).

\begin{comment}
JF: Yet you do more than evaluating this three measures so this has to be updated. 
JF: Not sure we should use L3 above since it is being redefined as a normalized measure here!
\end{comment}

The focus of this paper is to evaluate different methods for predicting PPIs using the existing knowledge of the network of interactions, which is represented as an undirected graph. The traditional way is of approaching the problem is usually based on social networks analysis, more specifically on the triadic closure principle (TCP), that states that the more common shared friends that two people have, the more likely that they know each other. As shown by previous studies, the mentioned approach fails because it does not consider the structural and chemical properties
of the proteins \cite{Kovacs2019}.

\begin{comment}
JF: Not sure this is the place to introduce TCP. You are already stating what the paper will do. Sees like TCP, should be introduced before the L3 measure is introduces, as part of the history of past approaches. Also TCP is not referred to later on...
\end{comment}

@@

%% read throughout the document but edited up here. I prefer to read the results before editing the materials and methods section. 
\section{Materials and Methods}

@@

% cannot compile doc given the way images are being imported

%\begin{figure}[h]
%\caption{Methods Comparison for \emph{HI-II-14}}
%
%\noindent \centering{}\includegraphics[width=1\columnwidth]{\string"/home/ocin/Documents/02 Asignaturas/10 GraphAnalytics/Project1/hi-ii-14.txt\string".eps}
%\end{figure}
%
%\begin{figure}[h]
%\caption{Methods Comparison for \emph{HI-TESTED}}
%
%\noindent \centering{}\includegraphics[width=1\columnwidth]{\string"/home/ocin/Documents/02 Asignaturas/10 GraphAnalytics/Project1/hi-tested.txt\string".eps}
%\end{figure}
